Reformulating phylogenetic mixed models to improve flexibility and speed

### 2018-10-11 main.tex
minor typo in methods

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process. They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/interept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
\mli{make a diagram!}
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

@@

We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species.
For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
For single-site model, the full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + b_{obs}\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix}
\Bigg) \\
b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
\end{align}
where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), and four random effect parameters (phylogenetic random intercept, phylogenetic random slope, correlation between phylogenetic random slope and intercept and observation variance).  
Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).

We extend the simulation model by adding multiple sites where each site will have one observation per species. The multiple-site full model is the following: 
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + b_{obs}\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix}
\Bigg) \\
(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{sp_{int}} & \sigma_{sp_{int},phy_{slope}} \\ 
\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
\end{bmatrix}
\Bigg) \\
b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
\end{align}
The multiple-site full simulation model will have five additional random effects (predictor and intercept level random effect of species and their correlation, random intercept of site and random intercept of species site interaction) compared to the single-site full model.
Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
Random intercept of species and site interactions is an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.

>>>>

There are two general challenges in linking the evolutionary process to a phylogenetic comparative statistical framework.
% Phylogenetic signal : the magnitude of phylogenetic correlation
The first challenge is to identify and segregate phylogenetic signals \citep{blomberg2003testing} and other sources of variation that are present in residuals.  
% \bmb{prev sentence doesn't seem be quite the right lead-in to the rest of the paragraph? You go on to talk about the tip-variation/residual-variation confounding issue. Maybe replace with something talking more specifically about resolving tip variation?}
The classic phylogenetic regression assumes that phylogenetic correlation (PC) among the residuals from a regression between the traits arises because the residual variation evolves along the branches of the phylogeny according to a Brownian motion process; if the traits are normally distributed and observed without any additional error or within-species variation, this leads to Felsenstein's method of phylogenetically independent contrasts (PICs) \citep{felsenstein1985phylogenies}.
More recent approaches -- including phylogenetic generalized linear mixed models (PGLMM) \citep{ives2011generalized}, Pagel's $\lambda$ \citep{pagel1999inferring}, and Blomberg's $K$ \citep{blomberg2003testing} --- build upon Felsenstein's PICs by considering different response distributions and allowing extra parameters to account for additional complexities in the evolutionary process. They partition the phylogenetically correlated residual variation into (1) phylogenetically uncorrelated residual variation (observation error or tip variation) and (2) phylogenetic signal  \citep[biological/evolutionary process error:][]{hansen2012interpreting}.
However, when multiple observations (or repeated measures) of the same species (taxon) are being observed, this introduces another level of variation; in this case, phylogenetic signal and tip variation are now both part of the evolutionary process error (we will call this species/intercept level variation) and residual variation is associated with observation level variation (the standard statistical definition).
\mli{See shiny}
%\bmb{This is a good point, well expressed. Does PGLMM lump tip variation and within-species variation, or is this primarily an issue with Blomberg/Pagel? Thinking about Boettiger's critique \cite{boettiger2013is} (don't know if it's been peer-reviewed, we could ask him), which might be resolved if we think explicitly about within-species variation \ldots
%\mli{I think it is more like our simulation test. In Boettiger's example, it is treating sister taxa as different species. So it is using a phylogenetic tree of 8 species instead of 4.}
Despite the addition of multiple observations per tip taxon (species) being statistically straightforward, it is rarely seen: existing platforms are unable to fit multiple observations, or unbalanced design data, or both.
% \bmb{OK, but we should be prepared to support this. Could we construct a table of existing approaches/platforms with their capabilities?}

@@

We generated test data based on the simple mixed model formulation \ref{eq:glmm1}, \ref{eq:glmm2}, \ref{eq:glmm3} with a single response variable $\bY$ and a single predictor continuous variable $\bX$ for $n$ (25, 50, and 100) species.
For wider range of comparisons, we simulate one observation ($X_i,Y_i$) per species.
For single-site model, the full simulation model is the following:
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}}) + (\beta_1 + b_{phy_{slope}}) X + b_{obs}\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix}
\Bigg) \\
b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
\end{align}
where it contains two fixed effect parameters ($\beta_0$ and $\beta_1$), three random effect parameters (phylogenetic random intercept, phylogenetic random slope and correlation between phylogenetic random slope and intercept) and observation/residual variance.  
Predictor and intercept level random variation/effect of species are not applicable in this simulation setting with single observation per species because it will confound with the observational level variation (residuals).

We extend the simulation model by adding multiple sites where each site will have one observation per species. The multiple-site full model is the following: 
\begin{align}
\bY & = (\beta_0 + b_{phy_{int}} + b_{sp_{int}} + b_{site}) + (\beta_1 + b_{phy_{slope}} + b_{sp_{slope}}) X + b_{sp:site} + b_{obs}\\
(b_{phy_{int}}, b_{phy_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{phy_{int}} & \sigma_{phy_{int},phy_{slope}} \\ 
\sigma_{phy_{int},phy_{slope}} & \sigma^2_{phy_{slope}}
\end{bmatrix}
\Bigg) \\
(b_{sp_{int}}, b_{sp_{slope}}) & \sim \textrm{MVN} \Bigg( 0, \begin{bmatrix}
\sigma^2_{sp_{int}} & \sigma_{sp_{int},sp_{slope}} \\ 
\sigma_{sp_{int},sp_{slope}} & \sigma^2_{sp_{slope}}
\end{bmatrix}
\Bigg) \\
b_{site} & \sim \textrm{MVN} ( 0 , \Sigma_{site}) \\
b_{sp:site} & \sim \textrm{MVN} (0, \textrm{Kron}(\bf{I}_{site}, \sigma^2_{phy})) \\
b_{obs} & \sim \textrm{MVN} ( 0 , \Sigma)
\end{align}
The multiple-site full simulation model will have five additional random effects (predictor and intercept level random effect of species and their correlation, random intercept of site and random intercept of species site interaction) compared to the single-site full model.
Predictor and intercept level random effect of species are now applicable in the multiple-site model because there are multiple (one in every site) observations per species. 
Random intercept of species and site interactions is an equivalent version of nested random effects or phylogenetic attraction described in \cite{helmus2007separating} with the main random effects.
